Translation of B Implementations to the LLVM: Specification

### 2014-03-19 DOCUMENTATION/specification.tex
Specification: refactorings, small corrections, reworked section on data.

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[1]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
\newcommand{\SBPos}[0]{\ensuremath{\sf\varphi}}
\newcommand{\SBType}[0]{\ensuremath{\sf\tau}}

@@

	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{StructItem} }\\

@@

	\multicolumn{3}{c}{\B{StructItem}\alt\B{ArrayItem} } \\

@@

\begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{StructType} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data struct : \B{StructType}} \\
        \hline
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
        \hline
	\B{base} & : &  \B{Name} \\
	\B{element} & : & seq \B{Name}\\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record struct : \B{Record}} \\
        \hline
	\B{elements} & : & seq \B{Name}\\
	\B{values} & : & seq \B{Exp}\\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Exp}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).  Sequence and optional attributes are denoted 
      ``seq'' and ``opt'',respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\newcommand{\myind}{\hspace*{2em}}
\framebox{
\begin{minipage}{.9\textwidth}
\noindent\textbf{typedef section:}
Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
the data space of the module: \\
\myind \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
In case the data space of a module is empty, this section is empty and the
type is not defined. \\
\\
\noindent\textbf{interface section:} 
Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
defined):\\
\myind \llvm{\%M\$ref\$ = type \%M\$state\$*} \\
The interface also contains declarations of one LLVM functions for each
operation of the module. \\
\myind \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
If the data space of the module is not empty, a function definition
corresponding to the initialisation is also emitted: \\
\myind \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})} \\
\\
\noindent\textbf{implementation section:}
Defines the functions declared in the interface (only for developed modules): \\
\myind \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
\myind \myind \llvm{\ListOf{\nt{block}}} \\
\myind \myind \llvm{exit: ret void} \\
\myind \llvm{\}}
\end{minipage}
}
\end{center}
\caption{Summary of the different sections and the pattern of LLVM code
  composing them.}
\label{fig:summary}
\end{figure}

@@

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name.  Also, we define
additional auxiliary functions responsible for producing such identifiers:
\begin{align*}
\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
\stateref(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$ref\$"} & \text{ (the type pointer to state)} \\
\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
\end{align*}
We assume an unlimited supply of labels and local variables: we use $\lbl$ to
return a fresh LLVM label and $\variable$ for a fresh local variable name. Also, let us define:
\begin{itemize}
\item $\developed(\B{n}) \isdef \B{n.impl} \neq \bot$ to test if a machine is
  developed.
\item $\stateless(\B{n}) \isdef \B{n.vars} =  \bot \land \forall i \in \B{n.impo}, \stateless(i)$ to test if a machine is stateless.
\item $\importedmodules(\B{n}) \isdef \{ \B{i.mach} \mid \B{i} \in \B{n.impo}
  \}$ to get the set of machines directly imported from implementation $\B{n}$.
\item $\importedinstances(\B{n}) \isdef \B{n.impo}$ to get the set of machines
  instances directly imported from implementation $\B{n}$.
\item $\trimportedmodules(\B{n}) \isdef \ldots$ to get the set of machines
transitively imported from implementation $\B{n}$.
\item $\trimportedinstances(\B{n}) \isdef \ldots$ to get the set of machines
instances transitively imported from implementation $\B{n}$.
\item $\idx(\B{n}) \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
  \B{n.root.vari})$ to get, from a given concrete variable $\B{n}$, the position
  of the corresponding element in the structure representing the state of the
  implementation. Here where $\mathit{index}(e, s)$ is a function that returns
  the position of $e$ in sequence $s$ and $\conc$ is the concatenation operator
  for sequences.
\end{itemize}

@@

The function responsible for executing an operation \B{n} has as parameter a
reference to the instance of the corresponding module, if it is stateful, and
one parameter for each input and output of the corresponding operation:
\begin{align*}
  \trad{\B{n}}{\B{oper}}_{\text{interface}} \isdef
  &
\LET t = \bot \triangleleft \stateless(\B{n.root}) \triangleright \stateref(\B{n.root}) \IN \\
& \llvm{"declare void \PH{\Global(\B{n})}
( \PH{t},
  \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}_{\text{decl}}},
  \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}_{\text{decl}}}) \nl"} \\
  \trad{\B{n}}{\B{inp}}_{\text{decl}} \isdef & \trad{\B{n.type}}{\B{Type}} \\
  \trad{\B{n}}{\B{out}}_{\text{decl}} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
\end{align*}
It is noteworthy that the outputs of B operations are encoded as pointer
parameters.
\subsection{The implementation section}

The implementation section is composed of the definitions for each function
in the corresponding interface:
\begin{align*}
  \trad{\B{n}}{\B{Mach}}_{\text{implementation}} \isdef
  & \LET \B{i} = \B{n.impl} = \bot \IN\\
  & \quad \IF \B{i} = \bot \THEN \llvm{""} \\
  & \quad \ELSE \\
  & \quad \quad \llvm{"\PH{\trad{\B{i}}{\B{init}}}} \\
  & \quad \quad \llvm{\PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
\end{align*}
Code generation for operations and the initialization is described in
sec.~\ref{sec:control}.

@@

The translation presented in this paper is currently limited to integers and
booleans. In the case of integers, the translation is based on the supposition
that the B0 definition of \B{INT} is the range $-2^{31}.. 2^{31}-1$. We have
$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}$ and $\trad{INT}{\B{Type}} \isdef
\llvm{i32}$.

B constants shall be translated to constants in the LLVM program. The
translation of scalar constants consists in substituting them with their value.
%TODO Look the following text. 
The translation of non-scalar constants will be defined in a next version of
this document. The definition of the translation for expressions is
described next in section~\ref{sec:expr}.

@@

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

@@

The code responsible for evaluating a comparison \B{r} is defined by the
following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
in the fresh variable $v$.
\begin{align*}
\lefteqn{\trad{r}{\B{Comp}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \IN \LET v = \variable \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
\end{align*}

\paragraph{Arithmetic operations} have a similar
structure and interpretation as comparisons. Here is the rule for binary
arithmetic operations:
\begin{align*}
\lefteqn{\trad{r}{\B{Term}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \IN \LET v = \variable \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = i32 \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v \sep \llvm{i32}
\end{align*}
Non-binary applications of associative operators are handled as nested
applications of binary applications. Next is the translation of the unary
\B{succ} operator:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t = \trad{n.arg}{\B{expr}} \IN  \\
  & \quad \LET w = \variable \IN \\
  & \quad \quad \llvm{"\PH{p}} \\
  & \quad \quad \llvm{\PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

@@

\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
following rule handles the case of conjunctions. Note that it produces code with
a ``short-cut'' when the evaluation of the first argument yields false:
\begin{align*}
\begin{split}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \lbl \IN} \\
  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
\end{split}
\end{align*}
The definition of the translation for disjunctions is similar and is omitted.

>>>>

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\DeclareMathOperator{\BE}{\hat{=}}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[1]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
\newcommand{\self}[0]{\llvm{\%self\$}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
\newcommand{\SBPos}[0]{\ensuremath{\sf\varphi}}
\newcommand{\SBType}[0]{\ensuremath{\sf\tau}}

@@

	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{RecField} }\\

@@

	\multicolumn{3}{c}{\B{RecField}\alt\B{ArrayItem} } \\

@@

\begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Structure} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Structure : \B{Structure}} \\
        \hline
        \B{elements} & : & seq \B{Field} \\
	\hline
  	\multicolumn{3}{|c|}{Structure field: \B{Field}} \\
        \hline
        \B{id} & : & \B{Name} \\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record: \B{Record}} \\
        \hline
	\B{type} & : & \B{Structure}\\
	\B{values} & : & seq \B{Exp}\\
	\hline
	\multicolumn{3}{|c|}{Record field : \B{RecField}} \\
        \hline
	\B{base} & : &  \B{name} \\
	\B{element} & : & seq \B{Name} \\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Exp}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).  Sequence and optional attributes are denoted 
      ``seq'' and ``opt'',respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\newcommand{\myind}{\hspace*{2em}}
\framebox{
\begin{minipage}{.9\textwidth}
\noindent\textbf{typedef section:}
Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
the data space of the module: \\
\myind \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
In case the data space of a module is empty, this section is empty and the
type is not defined. \\
\\
\noindent\textbf{interface section:} 
Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
defined):\\
\myind \llvm{\%M\$ref\$ = type \%M\$state\$*} \\
The interface also contains declarations of one LLVM functions for each
operation of the module. \\
\myind \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
If the data space of the module is not empty, a function definition
corresponding to the initialisation is also emitted: \\
\myind \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})} \\
\\
\noindent\textbf{implementation section:}
Defines the functions declared in the interface (only for developed modules): \\
\myind \llvm{define void @M\$op(\%M\$state\$* \self, \ListOf{\nt{param}}) \{} \\
\myind \myind \llvm{\ListOf{\nt{block}}} \\
\myind \myind \llvm{exit: ret void} \\
\myind \llvm{\}}
\end{minipage}
}
\end{center}
\caption{Summary of the different sections and the pattern of LLVM code
  composing them.}
\label{fig:summary}
\end{figure}

@@

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name.  Also, we define
additional auxiliary functions responsible for producing such identifiers:
\begin{align*}
\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
\stateref(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$ref\$"} & \text{ (the type pointer to state)} \\
\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)}
\end{align*}
We assume an unlimited supply of labels and local variables: we use $\lbl$ to
return a fresh LLVM label and $\variable$ for a fresh local variable name. Also, let us define:
\begin{itemize}
\item $\developed(\B{n}) \isdef \B{n.impl} \neq \bot$ to test if a machine is
  developed.
\item $\stateless(\B{n}) \isdef \B{n.vars} =  \bot \land \forall i \in \B{n.impo}, \stateless(i)$ to test if a machine is stateless.
\item $\importedmodules(\B{n}) \isdef \{ \B{i.mach} \mid \B{i} \in \B{n.impo}
  \}$ to get the set of machines directly imported from implementation $\B{n}$.
\item $\importedinstances(\B{n}) \isdef \B{n.impo}$ to get the set of machines
  instances directly imported from implementation $\B{n}$.
\item $\trimportedmodules(\B{n}) \isdef \ldots$ to get the set of machines
transitively imported from implementation $\B{n}$.
\item $\trimportedinstances(\B{n}) \isdef \ldots$ to get the set of machines
instances transitively imported from implementation $\B{n}$.
\item $\idx(\B{n}) \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
  \B{n.root.vari})$ to get, from a given concrete variable $\B{n}$, the position
  of the corresponding element in the structure representing the state of the
  implementation. Here where $\mathit{index}(e, s)$ is a function that returns
  the position of $e$ in sequence $s$ and $\conc$ is the concatenation operator
  for sequences.
\end{itemize}

@@

The function responsible for executing an operation \B{n} has as parameter a
reference to the instance of the corresponding module, if it is stateful, and
one parameter for each input and output of the corresponding operation:
\begin{align*}
  \trad{\B{n}}{\B{oper}}_{\text{interface}} \isdef
  &
\LET t \BE \bot \triangleleft \stateless(\B{n.root}) \triangleright \stateref(\B{n.root}) \IN \\
& \llvm{"declare void \PH{\Global(\B{n})}
( \PH{t},
  \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}_{\text{decl}}},
  \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}_{\text{decl}}}) \nl"} \\
  \trad{\B{n}}{\B{inp}}_{\text{decl}} \isdef & \trad{\B{n.type}}{\B{Type}} \\
  \trad{\B{n}}{\B{out}}_{\text{decl}} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
\end{align*}
It is noteworthy that the outputs of B operations are encoded as pointer
parameters.
\subsection{The implementation section}

The implementation section is composed of the definitions for each function
in the corresponding interface:
\begin{align*}
  \trad{\B{n}}{\B{Mach}}_{\text{implementation}} \isdef
  & \LET \B{i} \BE \B{n.impl} = \bot \IN\\
  & \quad \IF \B{i} = \bot \THEN \llvm{""} \\
  & \quad \ELSE \\
  & \quad \quad \llvm{"\PH{\trad{\B{i}}{\B{init}}}} \\
  & \quad \quad \llvm{\PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
\end{align*}
Code generation for operations and the initialization is described in
sec.~\ref{sec:control}.

@@

The translation presented in this paper handles the following concrete data:
\begin{description}
\item[Booleans] There is no Boolean type in LLVM, so we encode Booleans with
  one-bit integer:
$$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}.$$
\item[integers] Atelier-B offers three possible definitions for integers,
  corresponding to 16-bit, 32-bit (default) and 64-bit integers. The code
  generator has an option to specify which definition is employed, and we have:
$$\trad{INT}{\B{Type}} \isdef \llvm{i16} \mid \llvm{i32} \mid \llvm{i64}.$$
\item[enumerated sets] Again, enumerations are encoded as integer types, and the
  width of the type depends on the size of the enumeration.
$$\trad{\{e1 ... en\}}{\B{Type}} \isdef \LET w \BE \log_2 \B{n} \IN \llvm{i\PH{w}}.$$
\item[arrays] In LLVM, an array type is defined by its size and the type of its
  elements, the main issues in encoding a B array type into LLVM being to
  determine the size of the array and multi-dimensionality.

  [[[ TODO ]]]

\item[structures] The structure types of B are naturally encoded as structure
  types in LLVM.

  [[[ TODO ]]]
\end{description}

The current specification of the code generator does not handle deferred sets.

@@

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t \BE \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p \BE \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \self, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

@@

The code responsible for evaluating a comparison \B{r} is defined by the
following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
in the fresh variable $v$.
\begin{align*}
\lefteqn{\trad{r}{\B{Comp}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 \BE \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 \BE \trad{r.arg2}{\B{Expr}}} \\
& \IN \LET v \BE \variable \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
\end{align*}

\paragraph{Arithmetic operations} have a similar
structure and interpretation as comparisons. Here is the rule for binary
arithmetic operations:
\begin{align*}
\lefteqn{\trad{r}{\B{Term}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 \BE \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 \BE \trad{r.arg2}{\B{Expr}}} \\
& \IN \LET v \BE \variable \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = i32 \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v \sep \llvm{i32}
\end{align*}
Non-binary applications of associative operators are handled as nested
applications of binary applications. Next is the translation of the unary
\B{succ} operator:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t = \trad{n.arg}{\B{expr}} \IN  \\
  & \quad \LET w \BE \variable \IN \\
  & \quad \quad \llvm{"\PH{p}} \\
  & \quad \quad \llvm{\PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

@@

\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
following rule handles the case of conjunctions. Note that it produces code with
a ``short-cut'' when the evaluation of the first argument yields false:
\begin{align*}
\begin{split}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell \BE \lbl \IN} \\
  & \LET p_1 \BE \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
  \AND p_2 \BE \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
\end{split}
\end{align*}
The definition of the translation for disjunctions is similar and is omitted.

\paragraph{$\lambda$-expressions}

[[[ TODO ]]]
